---
title: "final_project"
author: "Kevin Gilds"
date: "8/10/2018"
output: html_document
---

```{r setup, include=FALSE}
## install packages if necessary
list.of.packages <- c("janitor", "ggplot2", "dplyr", "lubridate", "zoo", "forcats", "ggthemes", "knitr", "rmarkdown")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## load packages
library(janitor)
library(ggplot2)
library(dplyr)
library(lubridate)
library(zoo)
library(forcats)
library(ggthemes)
library(knitr)
library(tidyr)


## set working directory
knitr::opts_knit$set(root.dir = '/cloud/project/final_project')
```

### The Data

The American Time Use Survey (ATUS) is a time-use survey of Americans, which is sponsored by the Bureau of Labor Statistics (BLS) and conducted by the U.S. Census Bureau. Respondents of the survey are asked to keep a diary for one day carefully recording the amount of time they spend on various activities including working, leisure, childcare, and household activities. The survey has been conducted every year since 2003.

Included in the data are main demographic variables such as respondents' age, sex, race, marital status, and education. The data also includes detailed income and employment information for each respondent. While there are some slight changes to the survey each year, the main questions asked stay the same. You can find the data dictionaries for each year on [https://www.bls.gov/tus/dictionaries.htm](https://www.bls.gov/tus/dictionaries.htm)


### Accessing the Data

There are multiple ways to access the ATUS data; however, for this project, you'll get the raw data directly from the source. The data for each year can be found at [https://www.bls.gov/tus/#data](https://www.bls.gov/tus/#data). Once there, there is an option of downloading a multi-year file, which includes data for all of the years the survey has been conducted, but **for the purposes of this project, let's just look at the data for 2016**. Under **Data Files**, click on `American Time Use Survey--2016 Microdata files`. 

You will be brought to a new screen. Scroll down to the section **2016 Basic ATUS Data Files**. Under this section, you'll want to **click to download** the following two files: `ATUS 2016 Activity summary file (zip)` and `ATUS-CPS 2016 file (zip)`. 

* `ATUS 2016 Activity summary file (zip)` contains information about the total time each ATUS respondent spent doing each activity listed in the survey. The activity data includes information such as activity codes, activity start and stop times, and locations.
* `ATUS-CPS 2016 file (zip)` contains information about each household member of all individuals selected to participate in the ATUS.

Once they've been downloaded, you'll need to **unzip the files**. Once unzipped, you will see the dataset in a number of different file formats including `.sas`, `.sps`, and `.dat` files. **We'll be working with the .dat files.**

### Loading the Data into R

Use the first approach explained above to download and access the ATUS data for 2016. Download the CPS and Activity Summary files in a folder and unzip them and within each folder upload the files ending in .dat to `data/raw_data` filder on RStudio.cloud. To load the data in, **run the code in the `atus-data` code chunk** to create an object called `atus.all`.

### Importing data

```{r atus-data, warning=FALSE}
atus.cps <- read.delim('data/raw_data/atuscps_2016.dat', sep=",")
atus.sum <- read.delim('data/raw_data/atussum_2016.dat', sep=",")
atus.all <- atus.sum %>%
  left_join(atus.cps %>% filter(TULINENO==1), by = c("TUCASEID"))
```

### Exploratory Analysis of Child Care Data

This code chunk contains code scripts to provide a structure of the data including counts and types of variables. 

```{r initial-exploration}
### Add Code Here
dim(atus.all)
str(atus.all)
glimpse(atus.all)
names(atus.all)

mean(atus.all$t120101, rm.na = TRUE)

```

This code chunk creates a childcare variable.

```{r creating-childcare-var}
  atus.all <- atus.all %>% 
  mutate(CHILDCARE = t030101 + t030102 + t030103 + t030104 + t030105 + t030106 + t030108 + t030109 + t030110 + t030110 + t030111 + t030112 + t030199)
    
```

The code chunk produces a density plot pertaining to the amount of time spent on childcare.

```{r childcare-density-plot}
ggplot(atus.all, aes(x = CHILDCARE)) +
    geom_density()
```

```{r gender-analysis}
atus.all %>% 
    group_by(TESEX) %>% 
    summarise(average = mean(CHILDCARE))
```

```{r replacing-na}
## replace -1 in the variable TRDPFTPT with NA.

atus.all <- atus.all %>%
  mutate(TRDPFTPT = ifelse(TRDPFTPT == -1, NA, TRDPFTPT))


glimpse(atus.all$TRDPFTPT)

table(atus.all$TRDPFTPT)

atus.all %>% tabyl(TRDPFTPT) %>%
  adorn_totals()
```

The code below produces exploratory plots and tables. 

```{r exploratory-analysis}
## add your exploratory analysis code here

ggplot(atus.all, aes(x= TEAGE, y = CHILDCARE)) + geom_jitter()



atus.all %>% tabyl(CHILDCARE, TEAGE)

ggplot(atus.all, aes(x= HEFAMINC,  y= CHILDCARE, group= HEFAMINC)) + geom_boxplot()

atus.all %>% tabyl(CHILDCARE, HEFAMINC)

ggplot(atus.all, aes(x=PEMARITL, y=CHILDCARE)) + geom_point ()

atus.all %>% tabyl(CHILDCARE, PEMARITL)

ggplot(atus.all, aes(x= TRDPFTPT, y = CHILDCARE)) + geom_jitter()

atus.all %>% tabyl(CHILDCARE, TRDPFTPT)



```

```{r}
atus.all %>%
  tabyl(TRYHHCHILD) %>%
  adorn_totals()
```
I used this link to try and figure out what this variable means. 
https://www.bls.gov/tus/freqvariables.pdf

```{r}
atus.all  <- atus.all %>%
  filter(TRYHHCHILD >0) %>%
  mutate(TRYHHCHILD = ifelse(TRYHHCHILD == -1, NA, TRYHHCHILD))
```



### Regression Analysis

The code chunk below produces a model to determine factors that correspond the the amount of time one spends on child care. 

```{r reg-analysis}
## add your regression analysis code here
reg_model <- lm(CHILDCARE ~TRDPFTPT + PEMARITL + TEAGE + TESEX + HEFAMINC + TRYHHCHILD, data = atus.all)
summary(reg_model)
```


### Exploratory Analysis of Age and Activities

The code chunk below takes the sum of the variables that belong to each major category. 


```{r activity-cats}
atus.wide <- atus.all %>%
    mutate(act01 = rowSums(atus.all[, grep("t01", names(atus.all))]),
           act02 = rowSums(atus.all[,grep("t02", names(atus.all))]),
           act03 = rowSums(atus.all[,grep("t03", names(atus.all))]),
           act04 = rowSums(atus.all[, grep("t04", names(atus.all))]),
           act05 = rowSums(atus.all[, grep("t05", names(atus.all))]),
           act06 = rowSums(atus.all[, grep("t06", names(atus.all))]),
           act07 = rowSums(atus.all[, grep("t07", names(atus.all))]),
           act08 = rowSums(atus.all[, grep("t08", names(atus.all))]),
           act09 = rowSums(atus.all[, grep("t09", names(atus.all))]),
           act10 = rowSums(atus.all[, grep("t10", names(atus.all))]),
           act11 = rowSums(atus.all[, grep("t11", names(atus.all))]),
           act12 = rowSums(atus.all[, grep("t12", names(atus.all))]),
           act13 = rowSums(atus.all[, grep("t13", names(atus.all))]),
           act14 = rowSums(atus.all[, grep("t14", names(atus.all))]),
           act15 = rowSums(atus.all[, grep("t15", names(atus.all))]),
           act16 = rowSums(atus.all[, grep("t16", names(atus.all))]),
           # act17 = , there is no category 17 in the data
           act18 = rowSums(atus.all[, grep("t18", names(atus.all))])) %>% 
    select(TUCASEID, TEAGE, HEFAMINC, starts_with("act"))
```

The code chunk below generates a table to  breakdown the most minutes on average spent by major activity

```{r}

atus.long %>% 
    group_by(ACTIVITY) %>% 
    summarise(AVGMINS = mean(MINS)) 
    
```


            

```{r wide-to-long}
atus.long_1 <- atus.wide %>% 
  gather("ACTIVITY", "MINS", 4:20)

#atus.long %>%
  #group_by(ACTIVITY) %>%
  #summarise(TotalMins = sum(MINS))
  # use code to convert the wide format to long.

act_5 <- atus.long %>%
  filter(ACTIVITY == "work activities") 

summary(act_5) #answer to quiz questions

atus.long$HEFAMINC <- factor(atus.long$HEFAMINC) 


atus.long$ACTIVITY <- factor(atus.long$ACTIVITY, labels = c("personal care", "household activities", "caretaking household-members", "caretaking non-household", "work activities", "education", "consumer purchases", "professional services", "household services", "government services", "eating-drinking", "socializing", "sports", "religious", "volunteer", "telephone", "traveling" ))


atus.long$HEFAMINC <- factor(atus.long$HEFAMINC) 


atus.long <- atus.long %>%
  mutate(HEFAMINC = recode(HEFAMINC,
                           
                           "1" = "Less than $10,000",
                           "2" = "Less than $10,000",
                           "3" = "Less than $10,000",
                           "4" = "$10,000 - $20,000",
                           "5" = "$10,000 - $20,000",
                           "6" = "$10,000 - $20,000",
                           "7" = "$20,000 - $35,000",
                           "8" = "$20,000 - $35,000",
                           "9" = "$20,000 - $35,000",
                           "10" = "$35,000 - $60,000",
                           "11" = "$35,000 - $60,000",
                           "12" = "$35,000 - $60,000",
                           "13" = "$60,000 - $100,000",
                           "14" = "$60,000 - $100,000",
                           "15" = "$100,000 and over",
                           "16" = "$100,000 and over"))
                           
                           
                           
levels(atus.long$HEFAMINC)


#atus.long$HEFAMINC <- fct_relevel(atus.long$HEFAMINC, "Less than $10,000")
#atus.long$HEFAMINC <- fct_relevel(atus.long$HEFAMINC, "$100,000 and over", after = Inf)
#levels(atus.long$HEFAMIC) <- rev(atus.long$HEFAMINC)
#mtcars$cyl2 <- factor(mtcars$cyl, labels = c("alpha", "beta", "gamma"))
str(atus.long)





```

The graph below provides multiple scatter plot to demonstrate the relationship between age and the activities. We can see personal activities, work activities and socializing vary by age. 
```{r age-activity}
atus.long %>% 
    group_by(ACTIVITY, TEAGE) %>% 
    summarise(AVGMINS = mean(MINS)) %>% 
    ggplot(.,aes(y= AVGMINS, x = TEAGE)) +
      geom_point() +
    facet_wrap(~ACTIVITY)
```

### Exploratory Analysis of Income and Activities

I split the data frame into two separate objects and merged them back together in order to get the percent of time spent on activities by income group. 

```{r}
mins_income <- atus.long %>%  # Create object to get minutes by Income
  group_by(ACTIVITY, HEFAMINC) %>%   # Group the object by activity and income
  mutate(AVGMINS = mean(MINS))%>%   # Get average of Minutes
  summarise(MIN_INCOME = sum(AVGMINS))  # Get the sume of the average minuts by income

  
mins_income

mins_activity <- atus.long %>% #create ojbect 
  group_by(ACTIVITY) %>% # group by major cateogry
  mutate(MIN_ACTIVITY = mean(MINS)) %>%
  summarise(MIN_ACTIVITY = sum(MIN_ACTIVITY)) # sum 




minutes_activity_income <- merge(mins_income, mins_activity, by= "ACTIVITY")

minutes_activity_income_tbl <- minutes_activity_income %>%
  mutate(p = MIN_INCOME/MIN_ACTIVITY) %>%
  mutate(ACTIVITY = reorder(ACTIVITY, MIN_INCOME)) 
  
minutes_activity_income_tbl  
  

```


The goal was to mimic a graph found [here](https://raw.githubusercontent.com/halhen/viz-pub/master/pastime-income/pastime.png). The graph below is arranged by the most frequent activities. 

```{r}

minutes_activity_income_tbl %>%
  mutate(p = MIN_INCOME/MIN_ACTIVITY) %>%
  mutate(ACTIVITY = reorder(ACTIVITY, MIN_ACTIVITY)) %>%
  ggplot(aes(reorder(ACTIVITY), x = ACTIVITY, y = p, fill = HEFAMINC, group =2)) +
  geom_bar(stat='identity', position="fill") +
  coord_flip() +
  scale_fill_brewer(palette="BrBG") +
  labs(title="Percent of Time Spent on Activities",
       subtitle="Time Use Survey of 10493 Americans",
       caption="Source: American Time Use Survey, 2016",
       fill="Income",
       y = "Percent of Time Spent on Activities") +
  theme_bw() +
	theme(panel.grid.major.y = element_blank(),
		  panel.grid.minor.y = element_blank(),
		  axis.text.x = element_blank() )


```


`


```{r save-plot}
## save the plot above

ggsave("figures/explanatory_figures/activity-income.png")
```